How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-04-12 analysis/paper/paper.Rmd
Finish major text editing

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artefact volumetric density, retouch frequency, composition of toolkits, and artefact raw materials. We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments. In addition, groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviours. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. A population increase followed after the appearance of stemmed points.

@@

# Keywords {-}
Human Behavioural Ecology, Stemmed points, Mobility, Site occupation, Landscape use, Summed Probability Distribution, Pleistocene, Korean Late Paleolithic

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artefacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artefacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies, and stone artefacts are our only window into these behaviours. Here we use a behavioural ecological framework to explore questions about the implications of the appearance of this new technology. We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea. These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].

The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Locality SDG 2, which dates to around ca. 32ka, contains fauna specimens, ostrich eggshell beads, and stone artefacts including blades, retouched flakes, cores, and debris along with well-preserved hearths. In Locality 9 dated to ca. 29ka, there is a small scatter of stone artefacts consisting of blades, Levallois flakes, cores and other retouched flakes. Excavations Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone]. Shuidonggou shows that although stemmed points are not present at all, a blade technology came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9). Advanced core and flake tools were likely to have been locally developed at 33 ka (SDG Locality 2 and 8), and at 10.8 ka microblade technology appears, but the origin is uncertain (SDG Locality 12) [@yi2014significance; @li2019history]. The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards the Last Glacial Maximum (LGM) [@yi2014significance].

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca. 36ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artefacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are Chert and Andesite [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

@@

The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged]. This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35ka BP [@seong2009emergence]. Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points to support the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.

@@

Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016], but modelling the land-use behaviours associated with these assemblages has rarely been undertaken. We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviours. This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We draw on behavioural ecological theory to model the effects of this technological change. Behavioural ecology theory offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook]. Specifically, we use the patch choice model, which is built around the marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal]. We model individual artefact types as a kind of resource patch. Foragers may find a more costly and complicated technology (such as stemmed points, relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal]. That is, they may move to a patch where they can stay longer (i.e. an artefact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].

@@

Second, we ask what changes in the way people used habitation sites and mobility were associated with the new technology? After the appearance of these new tools, did people tend to be more or less mobile? The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. For example, stone artefact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics]. Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station. On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].

@@

## Sites and dates {-}

After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyse the process of technological change. This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artefact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

## Assemblages {-}

The 33 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artefacts from our sample because of uncertainty about their typology. The stone artefacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections to obtain permission to study the materials.

## Statistical methods {-}

To answer the two research questions, we compiled information about artefact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages.

The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and serves as a proxy for the accumulation ratio of artefacts [@clark2017lithics; @clark2019landscapes]. We used these density values to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene. The retouch frequency is the number of retouched tools divided by the total number of artefacts in an assemblage. This value represents mobility and duration of site occupation based on the assumption that tool reduction is a tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes]. In other words, producing and carrying highly retouched tools will increase portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling]. Combining these two proxies, the artefact volumetric density and the retouch frequency, we infer whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].

We calculated toolkit composition by analysing tool and debitage frequencies to identify changes in the assemblages during the transition period. Toolkit composition helps us answer the question of how site occupation patterns changed before and after the appearance of new technology. The type and proportion of retouched tools and knapping products, including debris and hammerstones, could give information about settlement patterns based on a premise that more activities are conducted with a wider range of tool types for longer occupation [@buvit2014late; @garcia2017occupational; @centi2019bulb]. For instance, small sized assemblages with a few types of tools represent short duration of site occupation [@shott2010size; @buvit2014late] .

Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korea relevant to our target period. For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from the simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We applied information of site location (all open-sites) including elevation, latitude and longitude to the simulated data set to get the annual temperature of our research period and area. Besides the climate analysis, we analysed site elevation as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation. We mainly focused on the elevation pattern of the sites before and after the appearance of new technology and its relationship with the stemmed points.

We used radiocarbon dates as a reference for population level. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon], including the lack of excavated sites in certain periods, sample size, and calibration effects, this method may allow a first approximation of population fluctuations. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, Timpson et al. apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243) [@timpson2014reconstructing]. We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic. He found that the lithics became less complex, and point shape changed as the population level dropped. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers]. For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental proxies and technological change. We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene]. We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].

All analyses and visualization are performed in the R Environment [@team2017r]. Our R code and data are available on the following GitHub repository: <https://github.com/parkgayoung/koreapaleolithicmobilityoccupation> as well as on the Open Science Framework: <https://osf.io/yagd2/> to enable full reproducibility [@marwick2017computational].

@@

## Artefact volumetric density and retouch frequency {-}

>>>>

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artifacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials. We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments. In addition, groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviours. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. A population increase followed after the appearance of stemmed points.

@@

# Keywords {-}
Human Behavioral Ecology, Stemmed points, Mobility, Site occupation, Landscape use, Summed Probability Distribution, Pleistocene, Korean Late Paleolithic

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies, and stone artifacts are our only window into these behaviors. Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea. These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].

The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Locality SDG 2, which dates to around ca. 32 ka, contains fauna specimens, ostrich eggshell beads, and stone artifacts including blades, retouched flakes, cores, and debris along with well-preserved hearths. In Locality 9 dated to ca. 29 ka, there is a small scatter of stone artifacts consisting of blades, Levallois flakes, cores and other retouched flakes. Excavations Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone]. Shuidonggou shows that although stemmed points are not present at all, a blade technology came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9). Advanced core and flake tools were likely to have been locally developed at 33 ka (SDG Locality 2 and 8), and at 10.8 ka microblade technology appears, but the origin is uncertain (SDG Locality 12) [@yi2014significance; @li2019history]. The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards the Last Glacial Maximum (LGM) [@yi2014significance].

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca. 36 ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are Chert and Andesite [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

@@

The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged]. This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35 ka BP [@seong2009emergence]. Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points to support the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.

@@

Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focused on timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016], but modeling the land-use behaviors associated with these assemblages has rarely been undertaken. We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviors. This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We draw on behavioral ecological theory to model the effects of this technological change. Behavioral ecology theory offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook]. Specifically, we use the patch choice model, which is built around the marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal]. We model individual artifact types as a kind of resource patch. Foragers may find a more costly and complicated technology (such as stemmed points, relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal]. That is, they may move to a patch where they can stay longer (i.e. an artifact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].

@@

Second, we ask what changes in the way people used habitation sites and mobility were associated with the new technology? After the appearance of these new tools, did people tend to be more or less mobile? The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics]. Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station. On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].

@@

## Archaeological Sites {-}

After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

## Stone Artifact Assemblages {-}

The 33 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artifacts from our sample because of uncertainty about their typology. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections to obtain permission to study the materials directly.

## Methods for Analyzing Stone Artifact Assemblages {-}

To answer our two research questions, we compiled information about artifact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages.

Artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and serves as a proxy for the accumulation ratio of artifacts [@clark2017lithics; @clark2019landscapes]. We used these density values to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene. Retouch frequency is the number of retouched tools divided by the total number of artifacts in an assemblage. This value represents mobility and duration of site occupation based on the assumption that tool reduction is a tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes]. In other words, producing and carrying highly retouched tools increased tool portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling]. Combining these two proxies, the artifact volumetric density and the retouch frequency, we infer whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has a high artifact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while lower artifact densities with higher proportions of retouched pieces shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].

We calculated toolkit composition by analyzing tool and debitage frequencies to identify changes in the assemblages during the study period. Toolkit composition helps us answer the question of how site occupation patterns changed before and after the appearance of new technology. The type and proportion of retouched tools and knapping products, including debris and hammerstones, provide information about settlement patterns based on a premise that more activities are conducted with a wider range of tool types for longer occupation [@buvit2014late; @garcia2017occupational; @centi2019bulb]. For example, we interpret small sized assemblages with a low diversity of tools as representing a short duration of site occupation [@shott2010size; @buvit2014late].

Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while a small assemblage of exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

## Methods for Analyzing Environmental Context {-}

In addition to these data on human behaviour from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korea relevant to our study period. For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from a simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We extracted annual temperature values from the data generated by @beyer2020high for all site locations in our sample. In addition, we extracted site elevation values from a raster, as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation, and their relationship with the stemmed points.

## Methods for Analyzing Demographic Context {-}

We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g. site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing]. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243). We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic. He found that the lithics became less complex, and point shape changed as the population level dropped. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models. For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental proxies and technological change. We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyzes and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene]. We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].

## Reproducibility and Open Source Materials {-}

The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at http://doi.org/osf.io/yagd2/ to enable re-use of materials and improve reproducibility and transparency [@Marwick2017]. Also in this version-controlled compendium [@marwick2018packaging] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

## Artifact Volumetric Density and Retouch Frequency {-}
